PLANETESIMAL ACCRETION AT SHORT ORBITAL PERIODS

### 2022-10-26 main.tex
Changes into annulus simulations.
Look for "TRQ" comments.

After a long period of quiescence, the collection of embryos and
remaining planetesimals undergoes a large-scale instability
\citep{chambers98}.
As a consequence of the instability, the oligarchics are no longer
on isolated, stable orbits and coalesce to form Earth-sized planets'
through a series of extremely energetic, giant 
impacts \citep{kokubo02, raymond05, raymond06}.

@@

Given that systems of tightly-packed inner planets (STIPs) appear to
be a common outcome of planet formation, understanding exactly how
solids accumulate at short orbital periods is crucial. Although
gas-disk driven migration of the planets themselves is often invoked
to explain the observed architectures, we will focus on an in-situ
model in this paper. That is, once the planetesimals themselves form,
they largely stay in place and any subsequent large-scale movement of
the solids are the result of mutual gravitational interactions. The
focus of this work will be to understand how the outcome of the
planetesimal accretion process scales with orbital period by using a
high-powered N-body code to directly follow the growth and evolution
of the planetesimals across a wide range of orbital periods (1 to 100
days). In doing so, we will assess whether the typical ``isolation
mass'' initial conditions used in studies of terrestrial planet formation are actually appropriate for understanding STIPs. In a series of follow up papers, we plan to use these results to directly track the formation of full-sized planets from planetesimals and to understand how gravitational perturbations from a nearby gas giant might affect the outcome of this process.

@@

We begin our analysis by considering a disk of equal planetesimals
with radius $r_{pl}$, mass $m_{pl}$ and surface density
$\Sigma_{pl}$. The collision rate in the vicinity of an orbit defined
by Keplerian frequency $\Omega$ can be written as $n \Gamma v$, where
$n = \Sigma_{pl} \Omega / 2 m_{pl} v$ (where we have assumed that the scale height of the planetesimal disk goes as $\Omega/(2v)$. $\Gamma$ describes the effective
collision cross section and $v$ is the typical encounter velocity
between planetesimals.
For a swarm of planetesimals on randomly oriented orbits, $v$ is typically
taken to the rms velocity, which can be related to the eccentricity and inclination distribution $(e, i)$ in the following way \citep{lissauer93}:
\begin{equation}\label{eq:ecc_vel}
	\left< v^{2} \right>^{1/2} = \left( \frac{5}{4} \left< e^{2} \right>^{1/2} + \left< i^{2} \right>^{1/2}  \right) v_{k}.
\end{equation}
Assuming that every collision results in a perfect merger, the growth rate of a planetesimal is given by
\begin{equation}\label{eq:growth}
	\frac{dM}{dt} = \frac{\Sigma \Omega}{2 m_{pl}} \Gamma.
\end{equation}

In the case where the collision cross section, depends only
on the physical size of the planetesimals, the growth scales linearly
with mass and the mass distribution is expected to evolve in an
``orderly'' fashion. However, bodies larger than $\sim 100$ km in size are expected to exert a significant gravitational force on each other during encounters and the collision cross section depends on both the size of the bodies and their encounter velocities. In this case, $\Gamma = \Gamma_{geo} \left( 1 + v_{esc}^2 / v^2 \right)$ \citep{safronov69}, where $v_{esc}$ is the escape velocity from the two bodies at the point of contact.

In the limit that $v_{esc} \gg v$, it can be shown that $M \propto
M^{4/3}$, which implies a runaway scenario, in which growth
accelerates with mass. This mode of growth was confirmed with N-body
simulations by \citet{kokubo96} and appears necessary to construct
protoplanets within the lifetime of a protoplanetary disk. Due to the
velocity dependence of the gravitational focusing effect, it is not clear how ubiquitous this mode of growth is. In particular, encounter velocities at short orbital periods will be rather large (because $v \sim v_{k}$) and the $v_{esc} \gg v$ condition may not always be satisfied. The effect that a dynamically hot disk has on runway growth will be examined in detail in section \ref{sec:narrow}.

@@

The picture described above relies upon an important assumption, which is that the mass distribution evolves slow enough for gravitational stirring to maintain energy equipartition. In other words, the relaxation timescale must remain short relative to the growth timescale. For typical conditions near the terrestrial planet forming region of the solar system, this timescale condition is satisfied. Due to the steep dependence of the relaxation time on encounter velocity, this condition can easily be violated at shorter orbital periods.

In figure \ref{fig:timescales}, we show ratio between the relaxation
and collision timescale for a population of equal-mass planetesimals
as a function of orbital period. Here, the encounter velocity is
described by equation \ref{eq:ecc_vel}. For
simplicity, we assume that $\left< e^2 \right>^{1/2} = 2\left< i^2
\right>^{1/2}$ \citep{ida93a} and that the eccentricity dispersion is
constant with orbital period. The eccentricity dispersion is described
in units scaled by the Hill factor $\left( m_{pl}/ 3 M_{\star}
\right)^{1/3}$ such that $e_{h} = 1$ corresponds to the boundary
between shear and dispersion dominated encounters. The horizontal
dashed line indicates where $t_{relax} = t_{coll}$. The timescale
criterion for oligarchic growth is only satisfied in regions where the
disk is sufficiently dynamically cold and the orbital period is
sufficiently long. In sections \ref{sec:narrow} and \ref{sec:fulldisk}
we will explore the behavior and outcome of planetesimal accretion in regions where this criterion is \textit{not} satisfied.

@@

In the formalism described above, the mass and velocity distribution of the bodies are both a function of time. Due to the interdependence of these quantities, it is not clear whether the timescales for gravitational scattering and growth will remain proportional as the oligarchs develop. In the case of many studies of planetesimal accretion in the solar system (cite examples), the $t_{relax} \ll t_{coll}$ condition must remain true, otherwise runaway growth would have continued until all of the planetesimals were consumed. However, it isn't immediately clear what will happen if the system begins in a state where $t_{coll} \ll t_{relax}$.

An insight into the expected behavior in this regime can be gained by
defining the dimensionless parameter $\alpha$, which is the ratio
between the physical size of a body and it's Hill radius, $r_{h}$:
\begin{equation}\label{eq:alpha}
	\alpha = \frac{r_{pl}}{r_{h}} = \frac{1}{a} \left( \frac{9 M_{\star}}{4 \pi \rho_{pl}} \right)^{1/3},
\end{equation}
where $a$ is the semimajor axis of the body and $\rho_{pl}$ is its bulk density. Assuming a fixed bulk density as bodies collide and grow, and that no large-scale migration occurs, the scaling of both $r_{pl}$ and $r_{h}$ with $m_{pl}^{1/3}$ means that $\alpha$ will be constant with time. For a composition of ice and rock, $\alpha$ is small for any populated region of the solar system ($\alpha \sim 10^{-2}$ for Earth and $\alpha \sim 10^{-4}$ in the Kuiper belt). As one moves close to the star $\alpha$ becomes larger than 1, which implies that the physical extent of a body exceeds its Hill sphere. (also mention that hill sphere size independent of velocity for small eccentricity)

@@

To answer these questions, we run a series of simulations in which a
narrow annulus of planetesimals orbits a star. The values of $\alpha$
and $\beta$ are varied individually. 4000 planetesimals with
individual masses of $5 \times 10^{23}$ g are placed with semimajor
axes randomly drawn between 0.95 and 1.05 AU about a 1 $M_{\odot}$
star. The argument of perihelion $\omega$, longitude of ascending node
$\Omega$, and mean anomaly M for each body is drawn from a uniform
distribution $\in [0, 2 \pi)$. The inclinations and eccentricities are drawn
from a Rayleigh distribution with
$\left< i^{2} \right> = 1/2 \left< e^{2} \right>$ \citep{ida93a}.

@@

In all cases, the simulations are evolved with a base timestep of 1.7
days, which corresponds to 3\% of an orbital dynamical time
$\sqrt{a^3/G M_{*}}$. Due to the vastly differing growth timescales in
each case, a simulation is stopped when the growth of the most massive
body flattens out. In figure \ref{fig:alpha_beta}, we show the a-e
distribution of bodies in the initial (blue) and final (orange)
snapshots from each of the 4 simulations. The size of the points
indicates the relative masses of the bodies. Only in the case of small
$\alpha$ does a residual population of dynamically hot planetesimals
develop. The lack of high eccentricity planetesimals in the large
$\alpha$ case suggests that most encounters result in accretion
instead of scattering. In the case of large $\beta$, the protoplanets
and remaining planetesimals end up in a dynamically cool state,
compared with the initial conditions. This is due to kinetic energy 
being lost as particles inelastically collide. One final point to note 
is the difference between eccentricities of the protoplanets in the 
large $\alpha$, large $\beta$ and the small $\alpha$,
large $\beta$ case. The dynamically cooler result of the latter case
is likely due to the dominant role that inelastic collisions play here.

@@

\begin{figure}
\begin{center}
    \includegraphics[width=0.5\textwidth]{figures/alpha_beta_mass.png}
    \caption{The final state of the mass distribution for the
      simulations described in section \ref{sec:narrow}. For small
      $\alpha$, a few embryos form alongside a power law tail of
      planetesimals. For larger values of $\alpha$, the mass
      distribution takes on a more uniform form. As in the previous figure, the initial choice of $\beta$ does not appear to have any meaningful impact on the end result.\label{fig:alpha_beta_mass}}
\end{center}
\end{figure}

>>>>

After a long period of quiescence, the collection of embryos and
remaining planetesimals undergoes a large-scale instability
\citep{chambers98}.
As a consequence of the instability, the oligarchics are no longer
on isolated, stable orbits and coalesce to form Earth-sized planets
through a series of extremely energetic, giant 
impacts \citep{kokubo02, raymond05, raymond06}.

@@

Given that systems of tightly-packed inner planets (STIPs) appear to
be a common outcome of planet formation, understanding exactly how
solids accumulate at short orbital periods is crucial. Although
gas-disk driven migration of the planets themselves is often invoked
to explain the observed architectures, we will focus on an in-situ
model in this paper. That is, once the planetesimals themselves form,
they largely stay in place, and any subsequent large-scale movement of
the solids are the result of mutual gravitational interactions. The
focus of this work will be to understand how the outcome of the
planetesimal accretion process scales with orbital period by using a
high-powered N-body code to directly follow the growth and evolution
of the planetesimals across a wide range of orbital periods (1 to 100
days). In doing so, we will assess whether the typical ``isolation
mass'' initial conditions used in studies of terrestrial planet formation are actually appropriate for understanding STIPs. In a series of follow up papers, we plan to use these results to directly track the formation of full-sized planets from planetesimals and to understand how gravitational perturbations from a nearby gas giant might affect the outcome of this process.

@@

We begin our analysis by considering a disk of equal planetesimals
with radius $r_{pl}$, mass $m_{pl}$ and surface density
$\Sigma_{pl}$. The collision rate in the vicinity of an orbit defined
by Keplerian frequency $\Omega$ can be written as $n \Gamma v$, where
$n = \Sigma_{pl} \Omega / 2 m_{pl} v$ (where we have assumed that the scale height of the planetesimal disk goes as $2v/\Omega$. $\Gamma$ describes the effective
collision cross section and $v$ is the typical encounter velocity
between planetesimals.
For a swarm of planetesimals on randomly oriented orbits, $v$ is typically
taken to the rms velocity, which can be related to the eccentricity and inclination distribution $(e, i)$ in the following way \citep{lissauer93}:
\begin{equation}\label{eq:ecc_vel}
	\left< v^{2} \right>^{1/2} = \left( \frac{5}{4} \left< e^{2} \right>^{1/2} + \left< i^{2} \right>^{1/2}  \right) v_{k}.
\end{equation}
Assuming that every collision results in a perfect merger, the growth rate of a planetesimal is given by
\begin{equation}\label{eq:growth}
	\frac{dM}{dt} = \frac{\Sigma \Omega}{2}} \Gamma.
\end{equation}

In the case where the collision cross section, depends only
% TRQ: perhaps "sub-linearly"?  I see dMdt ~ M^{2/3}
on the physical size of the planetesimals, the growth scales linearly
with mass and the mass distribution is expected to evolve in an
``orderly'' fashion. However, bodies larger than $\sim 100$ km in size are expected to exert a significant gravitational force on each other during encounters and the collision cross section depends on both the size of the bodies and their encounter velocities. In this case, $\Gamma = \Gamma_{geo} \left( 1 + v_{esc}^2 / v^2 \right)$ \citep{safronov69}, where $v_{esc}$ is the escape velocity from the two bodies at the point of contact.

In the limit that $v_{esc} \gg v$, it can be shown that $dM/dt \propto
M^{4/3}$, which implies a runaway scenario, in which growth
accelerates with mass. This mode of growth was confirmed with N-body
simulations by \citet{kokubo96} and appears necessary to construct
protoplanets within the lifetime of a protoplanetary disk. Due to the
velocity dependence of the gravitational focusing effect, it is not clear how ubiquitous this mode of growth is. In particular, encounter velocities at short orbital periods will be rather large (because $v \sim v_{k}$) and the $v_{esc} \gg v$ condition may not always be satisfied. The effect that a dynamically hot disk has on runway growth will be examined in detail in section \ref{sec:narrow}.

@@

The picture described above relies upon an important assumption, which
is that the mass distribution evolves slowly enough for gravitational stirring to maintain energy equipartition. In other words, the relaxation timescale must remain short relative to the growth timescale. For typical conditions near the terrestrial planet forming region of the solar system, this timescale condition is satisfied. Due to the steep dependence of the relaxation time on encounter velocity, this condition can easily be violated at shorter orbital periods.

In figure \ref{fig:timescales}, we show the ratio between the relaxation
and collision timescale for a population of equal-mass planetesimals
as a function of orbital period. Here, the encounter velocity is
described by equation \ref{eq:ecc_vel}. For
simplicity, we assume that $\left< e^2 \right>^{1/2} = 2\left< i^2
\right>^{1/2}$ \citep{ida93a} and that the eccentricity dispersion is
constant with orbital period. The eccentricity dispersion is described
in units scaled by the Hill factor $\left( m_{pl}/ 3 M_{\star}
\right)^{1/3}$ such that $e_{h} = 1$ corresponds to the boundary
between shear and dispersion dominated encounters. The horizontal
dashed line indicates where $t_{relax} = t_{coll}$. The timescale
criterion for oligarchic growth is only satisfied in regions where the
disk is sufficiently dynamically cold and the orbital period is
sufficiently long. In sections \ref{sec:narrow} and \ref{sec:fulldisk}
we will explore the behavior and outcome of planetesimal accretion in regions where this criterion is \textit{not} satisfied.

@@

In the formalism described above, the mass and velocity distribution
of the bodies are both a function of time. Due to the interdependence
of these quantities, it is not clear whether the timescales for
gravitational scattering and growth will remain proportional as the
oligarchs develop. In the case of many studies of planetesimal
accretion in the solar system
%TRQ: citations needed
{\bf(cite examples)}, the $t_{relax} \ll t_{coll}$ condition must remain true, otherwise runaway growth would have continued until all of the planetesimals were consumed. However, it isn't immediately clear what will happen if the system begins in a state where $t_{coll} \ll t_{relax}$.

An insight into the expected behavior in this regime can be gained by
defining the dimensionless parameter, $\alpha$, which is the ratio
between the physical size of a body and its Hill radius, $r_{h}$:
\begin{equation}\label{eq:alpha}
	\alpha = \frac{r_{pl}}{r_{h}} = \frac{1}{a} \left( \frac{9 M_{\star}}{4 \pi \rho_{pl}} \right)^{1/3},
\end{equation}
where $a$ is the semimajor axis of the body and $\rho_{pl}$ is its
bulk density. Assuming a fixed bulk density as bodies collide and
grow, and that no large-scale migration occurs, the scaling of both
$r_{pl}$ and $r_{h}$ with $m_{pl}^{1/3}$ means that $\alpha$ will be
constant with time. For a composition of ice and rock, $\alpha$ is
small for any populated region of the solar system ($\alpha \sim
10^{-2}$ for Earth and $\alpha \sim 10^{-4}$ in the Kuiper belt). As
one moves close to the star $\alpha$ becomes larger than 1, which
implies that the physical extent of a body exceeds its Hill sphere.
{\bf(also mention that hill sphere size independent of velocity for small eccentricity)}

@@

To answer these questions, we run a series of simulations in which a
narrow annulus of planetesimals orbits a star. The values of $\alpha$
and $\beta$ are varied individually. 4000 planetesimals with
individual masses of $5 \times 10^{23}$ g are placed with semimajor
axes randomly drawn between 0.95 and 1.05 AU about a 1 $M_{\odot}$
star. The argument of perihelion, $\omega$, longitude of ascending node,
$\Omega$, and mean anomaly, M, for each body is drawn from a uniform
distribution $\in [0, 2 \pi)$. The inclinations and eccentricities are drawn
from a Rayleigh distribution with
$\left< i^{2} \right> = 1/2 \left< e^{2} \right>$ \citep{ida93a}.

@@

In all cases, the simulations are evolved with a base timestep of 1.7
days, which corresponds to 3\% of an orbital dynamical time
$\sqrt{a^3/G M_{*}}$. Due to the vastly differing growth timescales in
each case, a simulation is stopped when the growth of the most massive
body flattens out. In figure \ref{fig:alpha_beta}, we show the a-e
distribution of bodies in the initial (blue) and final (orange)
snapshots from each of the 4 simulations. The size of the points
indicates the relative masses of the bodies. Only in the case of small
$\alpha$ does a residual population of dynamically hot planetesimals
develop. The lack of high eccentricity planetesimals in the large
$\alpha$ case suggests that most encounters result in accretion
instead of scattering. In the case of large $\beta$, the protoplanets
and remaining planetesimals end up in a dynamically cool state,
compared with the initial conditions. This is due to kinetic energy 
being lost as particles inelastically collide. One final point to note 
is the difference between eccentricities of the protoplanets in the 
large $\alpha$, large $\beta$ and the small $\alpha$,
%TRQ large alpha (former) is cooler!?!
large $\beta$ case. The dynamically cooler result of the latter case
is likely due to the dominant role that inelastic collisions play here.
%TRQ Moving figure 2 up in the paper would help keep this discussion straight.

@@

\begin{figure}
\begin{center}
    \includegraphics[width=0.5\textwidth]{figures/alpha_beta_mass.png}
    \caption{The final state of the mass distribution for the
      simulations described in section \ref{sec:narrow}. For small
      $\alpha$, a few embryos form alongside a power law tail of
      planetesimals. For larger values of $\alpha$, the mass
      distribution is more uniform.  As in the previous figure, the initial choice of $\beta$ does not appear to have any meaningful impact on the end result.\label{fig:alpha_beta_mass}}
\end{center}
\end{figure}
